% 
%            ,,,                                        
%          `7MM            _.o9                                
%            MM                                             
%  ,6"Yb.    MM  ,p6"bo   ,6"Yb.  M"""MMV  ,6"Yb.  `7Mb,od8 
% 8)   MM    MM 6M'  OO  8)   MM  '  AMV  8)   MM    MM' "' 
%  ,pm9MM    MM 8M        ,pm9MM    AMV    ,pm9MM    MM     
% 8M   MM    MM YM.    , 8M   MM   AMV  , 8M   MM    MM     
% `Moo9^Yo..JMML.YMbmd'  `Moo9^Yo.AMMmmmM `Moo9^Yo..JMML.   
%
%
% Neural Network Textbook - Abstract

\newpage
\thispagestyle{empty}

\begin{center}
    \Large\textbf{Abstract}
\end{center}

\vspace{1cm}

Neural networks have revolutionized the field of artificial intelligence, enabling breakthroughs in computer vision, natural language processing, speech recognition, and many other domains. This textbook provides a comprehensive introduction to neural networks, from fundamental concepts to state-of-the-art architectures.

The book is organized into eleven chapters, covering:

\begin{enumerate}
    \item \textbf{Mathematical Foundations:} Linear algebra, calculus, and probability theory essentials for understanding neural networks.
    
    \item \textbf{Activation Functions:} Comprehensive analysis of sigmoid, tanh, ReLU, and advanced activation functions with guidance on selection criteria.
    
    \item \textbf{Loss Functions:} Detailed exploration of cross-entropy, mean squared error, and task-specific loss functions.
    
    \item \textbf{Neural Network Architectures:} Deep Neural Networks (DNN), Convolutional Neural Networks (CNN), and Recurrent Neural Networks (RNN).
    
    \item \textbf{Core Operations:} Convolution, pooling, and normalization operations with mathematical rigor.
    
    \item \textbf{Advanced Mechanisms:} Skip connections and channel attention mechanisms including residual learning.
    
    \item \textbf{Optimization:} Gradient descent algorithms, optimizers (SGD, Adam, AdamW), and learning rate schedulers.
    
    \item \textbf{Classic Architectures:} From LeNet and AlexNet to ResNet, VGG, and the Transformer architecture.
    
    \item \textbf{Computational Analysis:} Parameter counting and FLOPs computation for network efficiency analysis.
    
    \item \textbf{Training and Frameworks:} Practical training procedures, hyperparameter tuning, and PyTorch implementation with CUDA concepts.
    
    \item \textbf{Dataset Processing:} Data collection, annotation, augmentation, and splitting strategies.
\end{enumerate}

Each chapter includes detailed mathematical derivations, practical examples, and references to seminal and recent research papers. Upon completing this textbook, readers will be equipped to design, implement, and train neural networks for various applications.

\vspace{2cm}

\textbf{Keywords:} Neural Networks, Deep Learning, Machine Learning, CNN, RNN, Transformer, Optimization, Backpropagation, PyTorch, Computer Vision, Natural Language Processing

\newpage
