\clearpage
\cleardoublepage

\chapter{Dataset Processing}

\section{Introduction}

The quality of a machine learning model is fundamentally limited by the quality of its training data. As the saying goes, ``garbage in, garbage out.'' This chapter covers the end-to-end pipeline of preparing data for neural network training: collection, annotation, splitting, augmentation, loading, normalization, and the increasingly important considerations of data quality and versioning.

While the techniques discussed apply broadly, we pay particular attention to the distinct requirements of different data modalities---images, text, and audio---since each presents unique processing challenges.

\section{Data Collection}

The foundation of any machine learning system is quality data. The dataset determines the ceiling of what your model can achieve.

\subsection{Data Sources}

Data can be obtained from multiple sources, each with its own quality and cost trade-offs:

\begin{itemize}
    \item \textbf{Public datasets:} Curated benchmark datasets (ImageNet, COCO, GLUE, SQuAD) provide standardized evaluation but may not match the target distribution.
    \item \textbf{Web scraping:} Automated collection from websites provides large volumes of data at low cost, but requires careful deduplication and quality filtering.
    \item \textbf{Sensors:} Cameras, microphones, medical imaging devices, accelerometers produce rich data but require careful calibration and preprocessing.
    \item \textbf{Human annotation:} Expert labeling provides high-quality ground truth but is expensive and slow. Crowdsourcing platforms (Amazon Mechanical Turk, Scale AI) trade annotation quality for scale and speed.
    \item \textbf{Synthetic data:} Data generated from simulations or generative models can augment limited real data, particularly valuable in medical imaging and robotics.
\end{itemize}

\subsection{Data Quality Criteria}

High-quality datasets share several properties:

\begin{enumerate}
    \item \textbf{Correctness:} Labels accurately reflect ground truth. Errors introduce bias and limit model performance ceiling.
    \item \textbf{Consistency:} Annotation guidelines are unambiguous, and annotators apply them uniformly.
    \item \textbf{Representativeness:} The dataset distribution matches the target deployment domain. A model trained on daylight images will fail on nighttime scenes.
    \item \textbf{Balancing:} Class frequencies and difficulty levels are appropriate for the task. Severe imbalance requires special handling.
    \item \textbf{Deduplication:} Near-duplicate samples inflate effective dataset size and cause train-test leakage.
\end{enumerate}

\section{Data Annotation}

Labels enable supervised learning. The annotation strategy depends on the task type.

\subsection{Annotation Types}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.8]
        % Classification
        \node[draw, rectangle, minimum width=2cm, minimum height=2cm, fill=blue!10] (cls) at (0,0) {};
        \node[below, font=\small] at (0, -1.3) {Classification};
        \node[right, font=\small] at (0.5, 0.3) {\textbf{Cat}};
        
        % Detection
        \node[draw, rectangle, minimum width=2cm, minimum height=2cm, fill=green!10] (det) at (4,0) {};
        \draw[red, thick] (3.3, 0.3) rectangle (4.3, 1.3);
        \draw[blue, thick] (3.7, -0.7) rectangle (4.7, 0.3);
        \node[below, font=\small] at (4, -1.3) {Object Detection};
        \node[right, red, font=\tiny] at (4.4, 0.8) {dog};
        \node[right, blue, font=\tiny] at (4.8, -0.2) {person};
        
        % Segmentation
        \node[draw, rectangle, minimum width=2cm, minimum height=2cm, fill=yellow!10] (seg) at (8,0) {};
        \fill[green!50, opacity=0.7] (7.3, 0.3) rectangle (8.3, 1.3);
        \fill[blue!50, opacity=0.7] (7.7, -0.7) rectangle (8.7, 0.3);
        \node[below, font=\small] at (8, -1.3) {Semantic Seg.};
    \end{tikzpicture}
    \caption{Annotation granularity: from image-level labels to pixel-level segmentation masks.}
    \label{fig:annotation_types}
\end{figure}

\begin{itemize}
    \item \textbf{Classification:} A single label per image (e.g., cat vs.\ dog). Simplest annotation task, prone to ambiguous cases.
    \item \textbf{Multi-label classification:} Multiple labels per image (e.g., ``beach, sunset, people''). Each label is an independent binary decision.
    \item \textbf{Object detection:} Bounding boxes with class labels. Each box specifies a region and its class.
    \item \textbf{Semantic segmentation:} Pixel-wise class labels. Every pixel is assigned a class (or ``background'').
    \item \textbf{Instance segmentation:} Distinguishing between individual objects of the same class (e.g., each person is a separate instance).
    \item \textbf{Keypoint detection:} Points of interest (facial landmarks, human body joints, vehicle detection points).
\end{itemize}

\subsection{Annotation Tools}

\begin{itemize}
    \item \textbf{LabelImg:} Open-source tool for bounding box annotation. Fast and lightweight.
    \item \textbf{CVAT:} Computer Vision Annotation Tool supports video, image, and text annotation with collaborative features.
    \item \textbf{Supervisely:} Full-featured platform with AI-assisted annotation (pre-labeling with existing models).
    \item \textbf{Amazon Mechanical Turk:} Crowdsourced human annotation at scale. Quality control via consensus/gold questions.
    \item \textbf{Label Studio:} Highly configurable open-source annotation tool supporting multiple data types.
\end{itemize}

\section{Data Splitting}

Proper data splitting is crucial for obtaining an unbiased estimate of model performance.

\subsection{Basic Splits}

\begin{itemize}
    \item \textbf{Training set:} The samples used to update model parameters. Typically 70--90\% of data.
    \item \textbf{Validation set:} Used for hyperparameter tuning and model selection. Never used for training.
    \item \textbf{Test set:} Used for final evaluation. Must never influence any training decisions.
\end{itemize}

\textbf{Typical ratios:}
\begin{itemize}
    \item Small datasets ($<$10K): 70/10/20 or 80/10/10 (train/val/test)
    \item Medium datasets (10K--1M): 80/10/10 or 85/5/10
    \item Large datasets ($>$1M): 98/1/1 may suffice, but ensure test set is large enough for reliable estimates
\end{itemize}

\begin{note}[Test Set Leakage]
Never use the test set for any decisions during training. This includes selecting hyperparameters, choosing architectures, or early stopping thresholds. Doing so leads to overfitting to the test set and unreliable performance estimates. The test set should be ``locked away'' until final evaluation.
\end{note}

\subsection{Stratified Splitting}

When class distributions are imbalanced, random splitting may produce unrepresentative splits. Stratified splitting maintains the class proportions in each split:

\begin{verbatim}
from sklearn.model_selection import train_test_split
X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.2, stratify=y
)
\end{verbatim}

For multi-label classification, stratification is more complex. One approach is to treat each label combination as a separate class, though this can lead to rare combinations. Alternatively, iterate through labels and balance each.

\subsection{Cross-Validation}

For robust evaluation with limited data, $K$-fold cross-validation provides more reliable estimates:

\begin{algorithm}[H]
\caption[K-Fold Cross-Validation]{$K$-Fold Cross-Validation}
\begin{enumerate}
    \item Shuffle data and split into $K$ equal folds
    \item For $i = 1, 2, \ldots, K$:
    \begin{enumerate}
        \item Train on folds $\{1, 2, \ldots, K\} \setminus \{i\}$
        \item Validate on fold $i$
    \end{enumerate}
    \item Report the average and standard deviation of the validation metrics
\end{enumerate}
\end{algorithm}

Common values: $K = 5$ (for moderate data) or $K = 10$ (for small data, larger variance reduction). Stratified $K$-fold maintains class proportions in each fold.

\section{Data Augmentation}

Data augmentation artificially increases dataset size by applying random transformations to existing samples, improving generalization without requiring additional data collection.

\subsection{Image Augmentations}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.85]
        % Original
        \node[draw, rectangle, minimum width=2cm, minimum height=2cm, fill=blue!20] (orig) at (0,0) {};
        \node[below, font=\small] at (0, -1.3) {Original};
        
        % Augmentations
        \node[draw, rectangle, fill=green!20] at (-3, -3) {};
        \node[below, font=\small] at (-3, -4.3) {Flipped};
        
        \node[draw, rectangle, fill=red!20] at (0, -3) {};
        \node[below, font=\small] at (0, -4.3) {Rotated};
        
        \node[draw, rectangle, fill=yellow!20] at (3, -3) {};
        \node[below, font=\small] at (3, -4.3) {Cropped};
        
        \node[draw, rectangle, fill=purple!20] at (6, -3) {};
        \node[below, font=\small] at (6, -4.3) {Color jitter};
        
        % Arrows
        \foreach \x in {-3, 0, 3, 6} {
            \draw[->, thick] (0, -1.0) -- (\x, -2.7);
        }
    \end{tikzpicture}
    \caption{Examples of image augmentations: horizontal flip, rotation, random crop, and color jitter.}
    \label{fig:augmentation}
\end{figure}

Common geometric augmentations:
\begin{itemize}
    \item \textbf{Rotation:} Random angle rotation (e.g., $\pm 15^\circ$). Should be within a range that preserves the label.
    \item \textbf{Flip:} Horizontal flip (right-left mirror). Safe for most classification tasks but not for text or asymmetric objects.
    \item \textbf{Random Crop:} Random crop to a smaller region, then resize to original size. Enforces scale invariance.
    \item \textbf{Scale:} Random scaling (zoom in/out). Same effect as random crop with different crop ratios.
    \item \textbf{Translation:} Random shifts in $x$ and $y$ directions. Equivalent to asymmetric cropping.
\end{itemize}

Color augmentations:
\begin{itemize}
    \item \textbf{Color jitter:} Random adjustments to brightness, contrast, saturation, and hue.
    \item \textbf{Gaussian noise:} Injection of pixel-wise Gaussian noise for robustness.
    \item \textbf{Cutout/Random Erasing:} Randomly masking rectangular regions. Forces the model to use distributed features.
    \item \textbf{Grayscale/RGB permutation:} For multi-channel images.
\end{itemize}

Mixup and CutMix augmentations blend samples together:

\subsection{Mixup: Convex Combinations of Samples}

Mixup \cite{zhang2018mixup} creates virtual training examples by linearly interpolating between pairs of samples:
\begin{align}
\tilde{\mathbf{x}} &= \lambda \mathbf{x}_i + (1 - \lambda)\mathbf{x}_j \\
\tilde{\mathbf{y}} &= \lambda \mathbf{y}_i + (1 - \lambda)\mathbf{y}_j
\end{align}
where $\lambda \sim \text{Beta}(\alpha, \alpha)$ for $\alpha \in (0, \infty)$. The mixing coefficient $\lambda$ controls the relative weight of each sample.

\textbf{Why Mixup works:}
\begin{enumerate}
    \item \textbf{Regularization:} The model sees smoothed decision boundaries, preventing overconfidence
    \item \textbf{Data cleaning:} Mixing noisy labels with clean ones dilutes label noise
    \item \textbf{Better calibration:} Predictions on mixed inputs encourage softer targets
    \item \textbf{Label smoothing effect:} For classification, $\tilde{y}$ is a soft label even when $y_i, y_j$ are one-hot
\end{enumerate}

For a $K$-class classification problem with one-hot labels $\mathbf{y}_i, \mathbf{y}_j \in \{0,1\}^K$, the mixed label $\tilde{\mathbf{y}}$ is a probability distribution over classes. The Mixup loss becomes:
\begin{equation}
\mathcal{L}_{\text{mixup}} = -\sum_{k=1}^{K} \tilde{y}_k \log \hat{y}_k = -\lambda \sum_k y_{i,k} \log \hat{y}_k - (1-\lambda) \sum_k y_{j,k} \log \hat{y}_k
\end{equation}

This is equivalent to minimizing the cross-entropy between the prediction and a mixture of the two one-hot labels.

\subsection{CutMix: Pasting Patches}

CutMix \cite{yun2019cutmix} pastes a rectangular region from one image into another instead of blending:
\begin{align}
\tilde{\mathbf{x}} &= \mathbf{x}_i \odot \mathbf{M} + \mathbf{x}_j \odot (1 - \mathbf{M}) \\
\tilde{\mathbf{y}} &= \lambda \mathbf{y}_i + (1 - \lambda)\mathbf{y}_j
\end{align}
where $\mathbf{M} \in \{0,1\}^{H \times W}$ is a binary mask and $\lambda = \frac{\text{area}(\mathbf{M})}{HW}$.

\textbf{CutMix vs.\ Mixup:} CutMix preserves the per-pixel nature of images (no blurring at the boundary) while still providing the regularization benefits of Mixup. The rectangular mask provides a clear contrast between regions, which can help models learn localized features.

\begin{code}
    \captionof{listing}{Mixup and CutMix in PyTorch}
    \label{code:mixup_cutmix}
\begin{lstlisting}[language=python]
import torch
import numpy as np

def mixup_data(x, y, alpha=1.0):
    """Returns mixed inputs, pairs of targets, and lambda."""
    if alpha > 0:
        lam = np.random.beta(alpha, alpha)
    else:
        lam = 1
    
    batch_size = x.size(0)
    index = torch.randperm(batch_size).to(x.device)
    
    mixed_x = lam * x + (1 - lam) * x[index, :]
    y_a, y_b = y, y[index]
    return mixed_x, y_a, y_b, lam

def mixup_criterion(criterion, pred, y_a, y_b, lam):
    return lam * criterion(pred, y_a) + (1 - lam) * criterion(pred, y_b)

def cutmix_data(x, y, alpha=1.0):
    """CutMix: Cuts and pastes patches between images."""
    if alpha > 0:
        lam = np.random.beta(alpha, alpha)
    else:
        lam = 1
    
    batch_size = x.size(0)
    index = torch.randperm(batch_size).to(x.device)
    
    # Get bounding box
    W = x.size(3); H = x.size(2)
    cut_rat = np.sqrt(1. - lam)
    cut_w = int(W * cut_rat); cut_h = int(H * cut_rat)
    
    cx = np.random.randint(W)
    cy = np.random.randint(H)
    bbx1 = np.clip(cx - cut_w // 2, 0, W)
    bby1 = np.clip(cy - cut_h // 2, 0, H)
    bbx2 = np.clip(cx + cut_w // 2, 0, W)
    bby2 = np.clip(cy + cut_h // 2, 0, H)
    
    x[:, :, bby1:bby2, bbx1:bbx2] = x[index, :, bby1:bby2, bbx1:bbx2]
    lam = 1 - ((bbx2 - bbx1) * (bby2 - bby1) / (W * H))
    
    y_a, y_b = y, y[index]
    return x, y_a, y_b, lam
\end{lstlisting}
\end{code}

\subsection{Advanced Augmentation Policies}

Rather than manually designing augmentations, automated methods search for optimal augmentation policies:

\begin{itemize}
    \item \textbf{AutoAugment} \cite{cubuk2019autoaugment}: Learned policies from CIFAR-10/CIFAR-100. Discovered policies transfer well to other datasets but are computationally expensive.
    \item \textbf{RandAugment} \cite{cubuk2020randaugment}: Simplified search space with just two parameters (magnitude, number of operations). Nearly matches AutoAugment with much less computation.
    \item \textbf{TrivialAugment Wide} \cite{muller2023torchvision}: No search needed. Applies a randomly selected augmentation at a random magnitude. Surprisingly effective and simple.
    \item \textbf{Test-Time Augmentation (TTA):} Apply augmentations at inference and average predictions. Provides free accuracy improvement (typically 1--2\%) at the cost of $K\times$ inference time.
\end{itemize}

\section{Text Data Processing}

Text data requires fundamentally different preprocessing than images, centered around the concept of tokenization.

\subsection{Tokenization}

Tokenization converts raw text strings into sequences of discrete tokens (word IDs or subword IDs) that the model can process.

\begin{table}[ht]
    \centering
    \caption{Tokenization Strategies}
    \begin{tabular}{L{3cm} L{4cm} L{4cm}}
        \toprule
        \textbf{Strategy} & \textbf{Example} & \textbf{Pros/Cons} \\
        \midrule
        Word-level & ``hello world'' $\to$ [12, 45] & Simple; large vocab, OOV issues \\
        Character-level & ``hello'' $\to$ [104, 101, 108, 108, 111] & Small vocab; long sequences, no semantics \\
        BPE & ``unhappiness'' $\to$ [``un'', ``happy'', ``ness''] & Good balance; learned from data \\
        WordPiece & BERT tokenizer & Similar to BPE; masked LM training \\
        SentencePiece & Unified tokenization & Language-agnostic; handles any text \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Byte Pair Encoding (BPE)}

BPE \cite{sennrich2016neural} is the most widely used subword tokenization algorithm in modern LLMs. It builds the vocabulary by iteratively merging the most frequent adjacent pairs of bytes/characters:

\begin{enumerate}
    \item Start with a vocabulary of all individual characters plus an end-of-word token
    \item Count all adjacent pairs in the corpus
    \item Merge the most frequent pair into a single token
    \item Repeat for $N$ merge operations (where $N$ is a hyperparameter)
\end{enumerate}

The vocabulary size is a key hyperparameter: larger vocabularies represent more words as single tokens (faster inference) but require more embeddings and have higher memory footprint.

For a vocabulary of size $V$, BPE requires $\mathcal{O}(V)$ space. Modern LLMs typically use $V = 30{,}000$--$100{,}000$ tokens.

\section{Audio Data Processing}

Audio data is typically processed as spectrograms or mel-spectrograms.

\subsection{Spectrogram Computation}

The short-time Fourier transform (STFT) converts a time-domain signal $x[t]$ into a time-frequency representation:

\begin{equation}
X[m, k] = \sum_{n=0}^{N-1} x[n] w[n] e^{-j 2\pi kn/N}
\end{equation}
where $w[n]$ is a window function (Hann or Hamming) and $N$ is the FFT size.

The \textbf{mel-spectrogram} maps frequencies to the mel scale, which approximates human auditory perception:
\begin{equation}
f_{\text{mel}} = 2595 \log_{10}\!\left(1 + \frac{f_{\text{Hz}}}{700}\right)
\end{equation}

\section{Data Normalization}

Normalization standardizes input features, which stabilizes training by ensuring all inputs are on similar scales.

\subsection{Image Normalization}

For RGB images, pixel values $x \in [0, 255]$ are typically normalized to $[0, 1]$ or standardized to zero mean and unit variance:

\begin{equation}
x_{\text{norm}} = \frac{x - \mu}{\sigma}
\end{equation}

ImageNet normalization parameters:
\begin{itemize}
    \item Mean: $[\,0.485,\, 0.456,\, 0.406\,]$
    \item Std: $[\,0.229,\, 0.224,\, 0.225\,]$
\end{itemize}

These values are computed from the ImageNet training set. For a custom dataset, compute dataset-specific statistics:

\begin{lstlisting}[language=python]
loader = DataLoader(dataset, batch_size=32)
mean = torch.zeros(3); std = torch.zeros(3)
for images, _ in loader:
    mean += images.mean(dim=[0, 2, 3])
    std += images.std(dim=[0, 2, 3])
mean /= len(loader.dataset)
std /= len(loader.dataset)
\end{lstlisting}

Note: When using pretrained models, always use the statistics the model was trained with, even if they are not optimal for your data.

\section{Data Loading}

Efficient data loading prevents GPU starvation and maximizes training throughput.

\begin{code}
    \captionof{listing}{Efficient Data Loading in PyTorch}
    \label{code:dataloader}
\begin{lstlisting}[language=python]
from torch.utils.data import DataLoader, Dataset

class ImageDataset(Dataset):
    def __init__(self, data, labels, transform=None):
        self.data = data
        self.labels = labels
        self.transform = transform
    
    def __len__(self):
        return len(self.data)
    
    def __getitem__(self, idx):
        image = self.data[idx]
        label = self.labels[idx]
        if self.transform:
            image = self.transform(image)
        return image, label

train_loader = DataLoader(
    train_dataset,
    batch_size=32,
    shuffle=True,         # Important for SGD convergence
    num_workers=4,        # Parallel data loading
    pin_memory=True,      # Faster GPU transfer (CUDA pinning)
    drop_last=True,       # Consistent batch sizes
    persistent_workers=True  # Keep workers alive between epochs
)

# Prefetch next epoch while current epoch trains
val_loader = DataLoader(
    val_dataset,
    batch_size=64,
    shuffle=False,         # No need to shuffle validation
    num_workers=4,
    pin_memory=True
)
\end{lstlisting}
\end{code}

\begin{properties}[DataLoader Optimization]
\begin{itemize}
    \item \textbf{num\_workers}: More workers = faster loading, but too many causes CPU-GPU contention. A good starting point is 4--8.
    \item \textbf{pin\_memory}: Pins CPU memory so transfers to GPU are faster (avoids an extra memory copy).
    \item \textbf{drop\_last}: Drops the last incomplete batch to avoid varying batch sizes (especially important for batch normalization).
    \item \textbf{persistent\_workers}: Keeps worker processes alive between epochs, reducing per-epoch startup overhead.
\end{itemize}
\end{properties}

\section{Handling Data Imbalance}

Class imbalance is ubiquitous in real-world datasets and can severely degrade model performance if not addressed.

\subsection{Imbalance Ratios}

Common imbalance ratios:
\begin{itemize}
    \item Mild (1:10 to 1:100): Standard medical imaging, anomaly detection
    \item Severe (1:100 to 1:1000): Fraud detection, rare disease diagnosis
    \item Extreme ($<$1:1000): Scientific applications (particle physics, astronomy)
\end{itemize}

\subsection{Techniques for Imbalanced Data}

\begin{enumerate}
    \item \textbf{Weighted loss function:} Assign higher loss weights to minority classes:
    \begin{equation}
    \mathcal{L}_{\text{weighted}} = -\sum_{k=1}^{K} w_k \cdot y_k \log \hat{y}_k
    \end{equation}
    where $w_k = N / (K \cdot N_k)$ and $N_k$ is the number of samples in class $k$.
    
    \item \textbf{Oversampling minority classes:} Replicate minority samples or generate synthetic samples (SMOTE for tabular data).
    
    \item \textbf{Undersampling majority classes:} Randomly remove samples from majority classes to balance the dataset.
    
    \item \textbf{Focal Loss} \cite{lin2017focal}: Down-weights well-classified examples, focusing training on hard examples:
    \begin{equation}
    \mathcal{L}_{\text{focal}} = -(1 - p_t)^\gamma \log p_t
    \end{equation}
    where $p_t$ is the model's predicted probability for the true class and $\gamma \geq 0$ controls the strength of down-weighting. See Chapter 3 for a detailed derivation.
\end{enumerate}

\section{Data Quality and Versioning}

As datasets grow in importance and complexity, systematic data management becomes essential.

\subsection{Data Versioning}

Data versioning tracks changes to datasets over time, enabling:
\begin{itemize}
    \item Reproducibility: Training with the exact same data at different points in time
    \item Lineage tracking: Understanding which model was trained on which version of the data
    \item Rollback: Reverting to previous data versions if issues are discovered
    \item Experiment comparison: Testing hypotheses about data quality
\end{itemize}

Tools like \textbf{DVC} (Data Version Control) extend Git to handle large files and ML pipelines:
\begin{verbatim}
dvc init
dvc add data/
git add data.dvc
git commit -m "Add dataset v1"
# Later, update and version:
dvc add data/
git add data.dvc
git commit -m "Update to dataset v2"
\end{verbatim}

DVC stores actual data in remote storage (S3, GCS, Azure Blob) while Git tracks metadata and hashes, enabling efficient collaboration.

\subsection{Data Provenance}

Understanding data provenance answers: Where did this sample come from? Who labeled it? When was it added?

Key provenance metadata:
\begin{itemize}
    \item Source URL or collection method
    \item Timestamp of collection/annotation
    \item Annotator ID and their qualifications
    \item Annotation quality scores (if using consensus annotations)
    \item Processing pipeline version (augmentations, filters applied)
\end{itemize}

This metadata is crucial for debugging model failures, auditing for bias, and satisfying regulatory requirements in high-stakes domains (healthcare, finance, legal).

\section{Evaluation Metrics}

While Chapter 3 covers loss functions, this section addresses task-specific evaluation metrics.

\subsection{Classification Metrics}

\begin{itemize}
    \item \textbf{Accuracy:} Fraction of correct predictions. Simple but misleading for imbalanced data.
    \item \textbf{Precision, Recall, F1:} For binary classification:
    \begin{align}
    \text{Precision} &= \frac{TP}{TP + FP} \quad \text{(of predicted positives, how many are correct?)} \\
    \text{Recall} &= \frac{TP}{TP + FN} \quad \text{(of actual positives, how many are found?)} \\
    F_1 &= \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
    \end{align}
    \item \textbf{Macro/Micro F1:} Macro-averages F1 per class then averages; micro-averages TP/FP/FN across classes then computes F1.
    \item \textbf{AUC-ROC:} Area under the ROC curve. Threshold-independent; measures ranking quality.
\end{itemize}

\subsection{Segmentation Metrics}

\begin{itemize}
    \item \textbf{Intersection over Union (IoU):} $\text{IoU} = \frac{|A \cap B|}{|A \cup B|}$
    \item \textbf{Dice Coefficient:} $\text{Dice} = \frac{2|A \cap B|}{|A| + |B|} = \frac{2 \cdot \text{IoU}}{1 + \text{IoU}}$
    \item \textbf{Pixel Accuracy:} Fraction of correctly classified pixels.
\end{itemize}

Note: IoU and Dice are related: $\text{Dice} = \frac{2 \cdot \text{IoU}}{1 + \text{IoU}}$, so optimizing one approximately optimizes the other.

\section{Chapter Summary}

\begin{enumerate}
    \item Data quality is the foundation of model performance; even the best architectures cannot overcome poor data
    \item Proper train/val/test splitting prevents leakage and provides unbiased performance estimates
    \item Cross-validation ($K$-fold) provides robust estimates when data is limited
    \item Data augmentation artificially expands the effective dataset through random transformations
    \item Mixup and CutMix create virtual training samples through convex combinations or patch pasting, providing strong regularization
    \item Text data requires tokenization (BPE, WordPiece) before model input
    \item Normalization stabilizes training by standardizing input scales
    \item Efficient data loading with multiple workers and pinned memory maximizes GPU utilization
    \item Class imbalance requires special handling through weighted losses or focal loss
    \item Data versioning and provenance tracking are essential for reproducible ML research and production systems
    \item Evaluation metrics should match the downstream task requirements; accuracy alone is insufficient for imbalanced or multi-class problems
\end{enumerate}

\section{Further Reading}

\begin{itemize}
    \item \cite{goodfellow2016deep} --- Chapter on Deep Learning, Section on Data Processing
    \item \cite{sennrich2016neural} --- Neural Machine Translation of Rare Words with Subword Units
    \item \cite{lin2017focal} --- Focal Loss for Dense Object Detection
    \item \cite{chollet2017deep} --- Deep Learning with Python, Chapter 4 on preprocessing pipelines
    \item \cite{zhang2018mixup} --- mixup: Beyond Empirical Risk Minimization
    \item \cite{cubuk2019autoaugment} --- AutoAugment: Learning Augmentation Policies from Data
    \item \cite{yun2019cutmix} --- CutMix: Regularization Strategy to Train Strong Classifiers with Localizable Features
    \item \cite{cubuk2020randaugment} --- RandAugment: Practical Automated Data Augmentation with a Reduced Search Space
    \item DVC Documentation: \url{https://dvc.org/} --- Data Version Control for ML projects
\end{itemize}

\section{Exercises}

\begin{enumerate}
    \item \textbf{Data Leakage:} Give two examples of data leakage in image classification or tabular ML pipelines. Why can leakage make validation accuracy misleadingly high?

    \item \textbf{Split Strategy:} You have 12,000 labeled examples and 5 classes with moderate imbalance. Design a train/validation/test split strategy and explain whether stratification is needed.

    \item \textbf{Mixup Labels:} If two images with one-hot labels $y_a$ and $y_b$ are combined with $\lambda = 0.3$, write the new target label used for training.

    \item \textbf{Augmentation Policies:} Compare AutoAugment, RandAugment, and TrivialAugment Wide. Which one is easiest to deploy when compute for policy search is limited?

    \item \textbf{Tokenization:} Explain why subword tokenization (BPE/WordPiece) usually works better than pure word-level tokenization for open-vocabulary NLP.

    \item \textbf{Input Pipeline Efficiency:} Why do multiple data-loader workers and pinned memory often improve GPU utilization? When might they fail to help?
\end{enumerate}

