\clearpage
\cleardoublepage

\chapter{Optimizers and Learning Rate Schedulers}

\section{Gradient Descent Optimization}

Training neural networks minimizes a loss function by updating parameters in the direction that reduces loss.

\subsection{Batch Gradient Descent}

Updates parameters using the gradient of the entire dataset:
\begin{equation}
\theta_{t+1} = \theta_t - \alpha \nabla_\theta \mathcal{L}(\theta_t)
\end{equation}

\begin{itemize}
    \item \textbf{Advantages:} Stable convergence, exact gradient
    \item \textbf{Disadvantages:} Slow for large datasets, memory intensive
\end{itemize}

\subsection{Stochastic Gradient Descent (SGD)}

Updates parameters using gradient of a single sample:
\begin{equation}
\theta_{t+1} = \theta_t - \alpha \nabla_\theta \mathcal{L}(\theta_t; \mathbf{x}^{(i_t)}, \mathbf{y}^{(i_t)})
\end{equation}

\begin{itemize}
    \item \textbf{Advantages:} Fast, escapes local minima, online learning
    \item \textbf{Disadvantages:} Noisy gradients, requires smaller learning rate
\end{itemize}

\subsection{Mini-Batch Gradient Descent}

Updates using a batch of $B$ samples:
\begin{equation}
\theta_{t+1} = \theta_t - \alpha \frac{1}{B}\sum_{i=1}^{B}\nabla_\theta \mathcal{L}(\theta_t; \mathbf{x}^{(i)}, \mathbf{y}^{(i)})
\end{equation}

This is the standard approach in deep learning.

\section{Gradient Descent with Momentum}

Momentum accelerates convergence by accumulating velocity:
\begin{align}
\mathbf{v}_{t+1} &= \beta \mathbf{v}_t + (1-\beta)\nabla_\theta \mathcal{L}(\theta_t) \\
\theta_{t+1} &= \theta_t - \alpha \mathbf{v}_{t+1}
\end{align}

\begin{itemize}
    \item $\beta \in [0, 1)$ is the momentum coefficient (typically 0.9)
    \item \textbf{Reduces oscillations} in steep directions
    \item \textbf{Accelerates convergence} in shallow directions
\end{itemize}

\begin{note}[Nesterov Accelerated Gradient]
Nesterov momentum looks ahead:
\begin{align}
\mathbf{v}_{t+1} &= \beta \mathbf{v}_t - \alpha \nabla_\theta \mathcal{L}(\theta_t + \beta\mathbf{v}_t) \\
\theta_{t+1} &= \theta_t + \mathbf{v}_{t+1}
\end{align}
This provides better convergence guarantees and often faster training.
\end{note}

\section{Adaptive Learning Rate Methods}

These methods adjust learning rates per-parameter based on gradient history.

\subsection{AdaGrad}

Adapts learning rate inversely proportional to accumulated squared gradients:
\begin{align}
\mathbf{g}_t &= \nabla_\theta \mathcal{L}(\theta_t) \\
\mathbf{r}_t &= \mathbf{r}_{t-1} + \mathbf{g}_t \odot \mathbf{g}_t \\
\theta_{t+1} &= \theta_t - \frac{\alpha}{\sqrt{\mathbf{r}_t + \epsilon}} \odot \mathbf{g}_t
\end{align}

\begin{itemize}
    \item \textbf{Good for sparse features}
    \item \textbf{Learning rate decreases} over time
    \item \textbf{May stop too early} in deep learning
\end{itemize}

\subsection{RMSProp}

Exponentially weighted moving average of squared gradients:
\begin{align}
\mathbf{g}_t &= \nabla_\theta \mathcal{L}(\theta_t) \\
\mathbf{r}_t &= \beta \mathbf{r}_{t-1} + (1-\beta)\mathbf{g}_t \odot \mathbf{g}_t \\
\theta_{t+1} &= \theta_t - \frac{\alpha}{\sqrt{\mathbf{r}_t + \epsilon}} \odot \mathbf{g}_t
\end{align}

\begin{itemize}
    \item \textbf{Addresses AdaGrad's diminishing learning rate}
    \item \textbf{Common default for RNNs}
    \item $\beta = 0.9$ is typical
\end{itemize}

\subsection{Adam (Adaptive Moment Estimation)}

Combines momentum and RMSProp:
\begin{align}
\mathbf{m}_t &= \beta_1 \mathbf{m}_{t-1} + (1-\beta_1)\mathbf{g}_t \quad \text{(first moment)} \\
\mathbf{v}_t &= \beta_2 \mathbf{v}_{t-1} + (1-\beta_2)\mathbf{g}_t^2 \quad \text{(second moment)} \\
\hat{\mathbf{m}}_t &= \frac{\mathbf{m}_t}{1-\beta_1^t} \quad \text{(bias correction)} \\
\hat{\mathbf{v}}_t &= \frac{\mathbf{v}_t}{1-\beta_2^t} \\
\theta_{t+1} &= \theta_t - \frac{\alpha \hat{\mathbf{m}}_t}{\sqrt{\hat{\mathbf{v}}_t} + \epsilon}
\end{align}

\begin{properties}[Adam Properties]
\begin{itemize}
    \item \textbf{Default optimizer} for most deep learning tasks
    \item \textbf{Robust to hyperparameters}
    \item \textbf{Default parameters:} $\alpha = 0.001, \beta_1 = 0.9, \beta_2 = 0.999, \epsilon = 10^{-8}$
    \item \textbf{Bias correction} for early time steps
\end{itemize}
\end{properties}

\begin{code}
    \captionof{listing}{Optimizers in PyTorch}
    \label{code:optimizers}
\begin{lstlisting}[language=python]
import torch.optim as optim

# SGD with momentum
optimizer = optim.SGD(model.parameters(), lr=0.01, momentum=0.9)

# SGD with Nesterov momentum
optimizer = optim.SGD(model.parameters(), lr=0.01, momentum=0.9, nesterov=True)

# Adam
optimizer = optim.Adam(model.parameters(), lr=0.001, betas=(0.9, 0.999))

# AdamW (decoupled weight decay)
optimizer = optim.AdamW(model.parameters(), lr=0.001, weight_decay=0.01)

# RMSProp
optimizer = optim.RMSprop(model.parameters(), lr=0.01, alpha=0.99)
\end{lstlisting}
\end{code}

\subsection{AdamW (Adam with Weight Decay)}

Decouples weight decay from gradient update:
\begin{equation}
\theta_{t+1} = \theta_t - \alpha \frac{\mathbf{m}_t}{\sqrt{\mathbf{v}_t} + \epsilon} - \alpha \lambda \theta_t
\end{equation}

\begin{itemize}
    \item \textbf{Proper L2 regularization} unlike original Adam
    \item \textbf{Better generalization} in many tasks
    \item \textbf{Standard in transformers}
\end{itemize}

\subsection{AdaDelta}

AdaDelta (Zeiler, 2012) addresses AdaGrad's monotonically decreasing learning rate by restricting the window of accumulated past gradients to a fixed size $w$ using an exponentially decaying average:
\begin{align}
\mathbf{r}_t &= \rho \mathbf{r}_{t-1} + (1 - \rho)\mathbf{g}_t^2 \\
\Delta\boldsymbol{\theta}_t &= -\frac{\sqrt{\mathbf{s}_{t-1} + \epsilon}}{\sqrt{\mathbf{r}_t + \epsilon}}\mathbf{g}_t \\
\mathbf{s}_t &= \rho \mathbf{s}_{t-1} + (1 - \rho)(\Delta\boldsymbol{\theta}_t)^2
\end{align}
where $\rho$ is a decay constant (typically 0.95).

\begin{properties}[AdaDelta Properties]
\begin{itemize}
    \item \textbf{No manual learning rate setting}: The numerator $\sqrt{\mathbf{s}_{t-1} + \epsilon}$ replaces $\alpha$, making the method largely hyperparameter-free
    \item \textbf{Robust to different gradient scales}: Adapts per-parameter using accumulated updates
    \item \textbf{Rarely used today}: Superseded by Adam and its variants
\end{itemize}
\end{properties}

\subsection{Unified View of Adaptive Methods}

\begin{table}[ht]
    \centering
    \caption{Adaptive Optimizer Family}
    \begin{tabular}{L{2cm} L{3cm} L{3cm} L{3.5cm}}
        \toprule
        \textbf{Method} & \textbf{1st moment} & \textbf{2nd moment} & \textbf{LR scaling} \\
        \midrule
        AdaGrad & No & $\mathbf{r}_t = \mathbf{r}_{t-1} + \mathbf{g}_t^2$ & $\alpha / \sqrt{\mathbf{r}_t}$ \\
        RMSProp & No & $\mathbf{r}_t = \rho\mathbf{r}_{t-1} + (1-\rho)\mathbf{g}_t^2$ & $\alpha / \sqrt{\mathbf{r}_t}$ \\
        AdaDelta & No & Same as RMSProp & $\sqrt{\mathbf{s}} / \sqrt{\mathbf{r}_t}$ \\
        Adam & $\mathbf{m}_t$ (bias corr.) & $\mathbf{v}_t$ (bias corr.) & $\alpha \hat{\mathbf{m}}_t / \sqrt{\hat{\mathbf{v}}_t}$ \\
        AdamW & Same as Adam & Same as Adam & + decoupled $\lambda\theta_t$ \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Convergence Rate Analysis}

\begin{theorem}[SGD Convergence Rate]
For a smooth ($L$-Lipschitz gradient) and $\mu$-strongly convex function $f$, gradient descent with learning rate $\alpha = 1/L$ satisfies:
\begin{equation}
f(\theta_t) - f(\theta^*) \leq \left(1 - \frac{\mu}{L}\right)^t \bigl(f(\theta_0) - f(\theta^*)\bigr)
\end{equation}
This is a \textbf{linear convergence rate} with rate $1 - \mu/L$.
\end{theorem}

\begin{properties}[Convergence Rates for Different Optimizers]
\begin{itemize}
    \item \textbf{Convex + smooth:} SGD achieves $O(1/T)$ rate; accelerated methods achieve $O(1/T^2)$
    \item \textbf{Non-convex:} All methods converge at $O(1/\sqrt{T})$ to a stationary point
    \item \textbf{Adam} is not guaranteed to converge in the non-convex setting without modification (Reddi et al., 2018); \textbf{AdamW} and \textbf{AMSGrad} fix this issue
    \item In practice, Adam and AdamW converge faster initially but SGD+Momentum often achieves better final generalization
\end{itemize}
\end{properties}

\section{Gradient Clipping}

Prevents exploding gradients. Two main strategies:

\subsection{Value Clipping}

Clip each gradient component individually:
\begin{equation}
\tilde{g}_i = \text{clip}(g_i, -c, c) = \max(\min(g_i, c), -c)
\end{equation}
where $c$ is the clip value (typically 0.5--5.0).

\subsection{Norm Clipping}

Scale the entire gradient vector if its norm exceeds a threshold:
\begin{equation}
\tilde{\mathbf{g}} = \begin{cases}
\mathbf{g} & \text{if } \|\mathbf{g}\| \leq c \\
\frac{c}{\|\mathbf{g}\|} \mathbf{g} & \text{if } \|\mathbf{g}\| > c
\end{cases}
\end{equation}
This preserves the gradient \textbf{direction} while capping its magnitude.

\begin{properties}[When to Use Gradient Clipping]
\begin{itemize}
    \item \textbf{RNNs and Transformers:} Essential to prevent exploding gradients through long sequences
    \item \textbf{Fine-tuning pretrained models:} Prevents destabilization from large gradient updates
    \item \textbf{Norm clipping} is generally preferred over value clipping (preserves gradient direction)
    \item Typical max norm: 1.0 (transformers) to 5.0 (RNNs)
    \item In PyTorch: \texttt{torch.nn.utils.clip\_grad\_norm\_}(model.parameters(), max\_norm=1.0)
\end{itemize}
\end{properties}

\section{Learning Rate Schedulers}

Learning rate schedules adjust the learning rate during training.

\subsection{Step Decay}

\begin{equation}
\alpha_t = \alpha_0 \cdot \gamma^{\lfloor t / T \rfloor}
\end{equation}
where $\gamma$ is the decay factor (e.g., 0.5) and $T$ is the period.

\subsection{Exponential Decay}

\begin{equation}
\alpha_t = \alpha_0 \cdot e^{-\lambda t}
\end{equation}

\subsection{Cosine Annealing}

\begin{equation}
\alpha_t = \alpha_{min} + \frac{1}{2}(\alpha_{max} - \alpha_{min})\left(1 + \cos\left(\frac{t\pi}{T}\right)\right)
\end{equation}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.8]
        \begin{axis}[
            xlabel=Epoch,
            ylabel=Learning Rate,
            width=10cm,
            height=5cm,
            legend pos=north east
        ]
            % Step decay
            \addplot[blue, const plot, mark=*] coordinates {
                (0, 0.01)
                (30, 0.005)
                (60, 0.0025)
                (90, 0.00125)
            };
            \addlegendentry{Step Decay}
            
            % Exponential
            \addplot[red, smooth] {0.01 * exp(-0.02*x)};
            \addlegendentry{Exponential}
            
            % Cosine
            \addplot[green, smooth] {(0.001 + 0.5*(0.01-0.001)*(1+cos(pi*x/100))};
            \addlegendentry{Cosine}
        \end{axis}
    \end{tikzpicture}
    \caption{Learning rate schedules}
    \label{fig:lr_schedules}
\end{figure}

\subsection{Warmup}

Gradually increase learning rate at the start:
\begin{equation}
\alpha_t = \alpha_{max} \cdot \frac{t}{T_{warmup}} \quad \text{for } t < T_{warmup}
\end{equation}

\begin{itemize}
    \item \textbf{Prevents large gradients} at the start
    \item \textbf{Important for transformers}
    \item \textbf{Common: 500-10000 warmup steps}
\end{itemize}

\subsection{Warmup + Cosine Decay}

The standard schedule for modern transformer training combines warmup with cosine annealing:
\begin{equation}
\alpha_t = \begin{cases}
\alpha_{max} \cdot \frac{t}{T_{warmup}} & \text{if } t < T_{warmup} \\
\alpha_{min} + \frac{1}{2}(\alpha_{max} - \alpha_{min})\left(1 + \cos\left(\frac{(t - T_{warmup})\pi}{T_{total} - T_{warmup}}\right)\right) & \text{otherwise}
\end{cases}
\end{equation}

This is the default in most transformer training recipes (BERT, GPT, ViT).

\subsection{ReduceLROnPlateau}

Monitors a metric (typically validation loss) and reduces the learning rate when improvement stalls:
\begin{equation}
\alpha_t = \alpha_{t-1} \cdot \gamma \quad \text{if } \text{no improvement for } p \text{ epochs}
\end{equation}
where $\gamma$ is the decay factor (typically 0.1) and $p$ is the patience (typically 5--10 epochs).

\begin{itemize}
    \item \textbf{Adaptive}: No need to predefine the total number of epochs
    \item \textbf{Useful when training time is unknown} (e.g., early stopping)
    \item Common in training CNNs for image classification
\end{itemize}

\subsection{OneCycleLR}

Proposed by Smith (2018), OneCycleLR uses a single cycle with three phases:
\begin{enumerate}
    \item \textbf{Increase} LR from $\alpha_{min}$ to $\alpha_{max}$ (first $\sim$30\% of steps)
    \item \textbf{Decrease} LR from $\alpha_{max}$ to $\alpha_{min}$ (next $\sim$60\% of steps)
    \item \textbf{Fine-tune}: Decrease LR further to $\alpha_{min}/100$ (final $\sim$10\% of steps)
\end{enumerate}
Optionally, the momentum follows an inverse pattern (high when LR is low).

\begin{properties}[OneCycleLR Benefits]
\begin{itemize}
    \item \textbf{Faster convergence}: Often achieves the same accuracy in fewer epochs
    \item \textbf{Regularization effect}: The high-to-low LR cycle acts as implicit regularization
    \item \textbf{No warmup needed}: Built into the schedule
    \item Popular in fastai and for transfer learning tasks
\end{itemize}
\end{properties}

\section{Optimizer Selection}

\begin{table}[ht]
    \centering
    \caption{Optimizer Selection Guide}
    \begin{tabular}{L{2.5cm} L{3cm} L{3.5cm}}
        \toprule
        \textbf{Optimizer} & \textbf{Good for} & \textbf{Notes} \\
        \midrule
        SGD + Momentum & Image classification & Best generalization \\
        Adam & General purpose & Default choice \\
        AdamW & Transformers, NLP & Best for fine-tuning \\
        RMSProp & RNNs & Good for recurrent models \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Chapter Summary}

\begin{enumerate}
    \item Gradient descent is the foundation of neural network optimization
    \item Momentum accelerates convergence by accumulating velocity
    \item Adaptive methods (AdaGrad, RMSProp, AdaDelta, Adam) adjust per-parameter learning rates
    \item Adam combines momentum and adaptive learning rates; AdamW adds proper weight decay
    \item Gradient clipping prevents exploding gradients (norm clipping preferred)
    \item Learning rate schedulers (step, cosine, plateau, OneCycleLR) improve convergence
    \item Warmup is important for transformer training
    \item SGD+Momentum often generalizes better; Adam converges faster
\end{enumerate}

\section{Further Reading}

\begin{itemize}
    \item Kingma, D. \& Ba, J. (2015). Adam: A method for stochastic optimization. \textit{ICLR}.
    \item Loshchilov, I. \& Hutter, F. (2019). Decoupled weight decay regularization. \textit{ICLR}.
    \item Zeiler, M. (2012). ADADELTA: An adaptive learning rate method. \textit{arXiv:1212.5701}.
    \item Reddi, S. et al. (2018). On the convergence of Adam and beyond. \textit{ICLR}.
    \item Smith, L. (2018). A disciplined approach to neural network hyper-parameters. \textit{NeurIPS Workshop}.
    \item Loshchilov, I. \& Hutter, F. (2017). SGDR: Stochastic gradient descent with warm restarts. \textit{ICLR}.
    \item Paszke, A. et al. (2019). PyTorch: An imperative style, high-performance deep learning library. \textit{NeurIPS}.
\end{itemize}
