\clearpage
\cleardoublepage

\chapter{Loss Functions}

\section{Introduction}

Loss functions (also called objective functions, cost functions, or error functions) are the mathematical measures that quantify how well a neural network's predictions match the true targets. The training process aims to minimize this loss by adjusting the network's parameters through gradient-based optimization.

The choice of loss function depends on:
\begin{itemize}
    \item The type of task (classification, regression, etc.)
    \item The distribution of the target variable
    \item The desired properties of the solution (sparsity, robustness, etc.)
    \item Computational considerations
\end{itemize}

\section{The Role of Loss Functions}

Given a dataset $\mathcal{D} = \{(\mathbf{x}_i, \mathbf{y}_i)\}_{i=1}^{N}$ and a model $f(\mathbf{x}; \theta)$, the empirical risk is:
\begin{equation}
\mathcal{R}(\theta) = \frac{1}{N}\sum_{i=1}^{N} \ell(f(\mathbf{x}_i; \theta), \mathbf{y}_i)
\end{equation}
where $\ell(\cdot, \cdot)$ is the loss function.

Training seeks to find:
\begin{equation}
\theta^* = \arg\min_{\theta} \mathcal{R}(\theta)
\end{equation}

The loss function should:
\begin{enumerate}
    \item Be differentiable (for gradient-based optimization)
    \item Penalize larger errors appropriately
    \item Encourage correct behavior for the specific task
    \item Have useful statistical interpretations
\end{enumerate}

\section{Regression Loss Functions}

Regression tasks predict continuous values. The choice of loss function encodes assumptions about the noise distribution and determines how the model handles outliers.

\subsection{Mean Squared Error (MSE)}

The Mean Squared Error (L2 loss) is the most common regression loss:
\begin{equation}
\mathcal{L}_{MSE} = \frac{1}{N}\sum_{i=1}^{N} (y_i - \hat{y}_i)^2 = \frac{1}{N}\sum_{i=1}^{N} e_i^2
\end{equation}
where $e_i = y_i - \hat{y}_i$ is the residual.

\subsubsection{Derivation: MSE as Maximum Likelihood under Gaussian Noise}

The statistical justification for MSE comes from maximum likelihood estimation. Suppose the data is generated by:
\begin{equation}
y_i = f(\mathbf{x}_i; \theta) + \epsilon_i, \quad \epsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2)
\end{equation}

The likelihood of a single observation is:
\begin{equation}
p(y_i \mid \mathbf{x}_i; \theta, \sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(y_i - \hat{y}_i)^2}{2\sigma^2}\right)
\end{equation}

The negative log-likelihood (NLL) for the full dataset:
\begin{align}
\mathcal{L}_{\text{NLL}} &= -\sum_{i=1}^{N} \log p(y_i \mid \mathbf{x}_i; \theta, \sigma^2) \\
&= -\sum_{i=1}^{N}\left[-\frac{1}{2}\log(2\pi\sigma^2) - \frac{(y_i - \hat{y}_i)^2}{2\sigma^2}\right] \\
&= \frac{N}{2}\log(2\pi\sigma^2) + \frac{1}{2\sigma^2}\sum_{i=1}^{N}(y_i - \hat{y}_i)^2
\end{align}

Since $\frac{N}{2}\log(2\pi\sigma^2)$ is a constant that does not depend on $\theta$:
\begin{equation}
\boxed{\arg\min_\theta \mathcal{L}_{\text{NLL}} = \arg\min_\theta \sum_{i=1}^{N}(y_i - \hat{y}_i)^2 = \arg\min_\theta \mathcal{L}_{MSE}}
\end{equation}

\textbf{Takeaway:} Minimizing MSE is equivalent to fitting a model under the assumption of additive, i.i.d.\ Gaussian noise. This is why MSE works well when errors are roughly Gaussian but fails when the noise distribution has heavy tails (e.g., Laplace noise).

\subsubsection{Gradient Analysis}

The gradient of MSE with respect to the prediction:
\begin{equation}
\frac{\partial \mathcal{L}_{MSE}}{\partial \hat{y}_i} = -\frac{2}{N}(y_i - \hat{y}_i) = \frac{2}{N}e_i
\end{equation}

The gradient is \textbf{linear in the error}---a double-edged sword:
\begin{itemize}
    \item \textbf{Near the optimum} ($e_i \approx 0$): Gradient is small, enabling precise convergence.
    \item \textbf{Far from the optimum} or \textbf{outliers} ($|e_i| \gg 0$): Gradient is very large, causing overshooting and excessive sensitivity to outliers.
\end{itemize}

\subsubsection{Connection to Variance and \texorpdfstring{$R^2$}{R^2}}

For a model $\hat{y}_i = f(\mathbf{x}_i)$, the total variance decomposition holds:
\begin{equation}
\underbrace{\sum_i (y_i - \bar{y})^2}_{\text{Total SS}} = \underbrace{\sum_i (\hat{y}_i - \bar{y})^2}_{\text{Explained SS}} + \underbrace{\sum_i (y_i - \hat{y}_i)^2}_{\text{Residual SS (MSE $\times N$)}}
\end{equation}

This gives the $R^2$ score: $R^2 = 1 - \frac{\text{MSE}}{\text{Var}(y)}$.

\subsection{Mean Absolute Error (MAE)}

The Mean Absolute Error (L1 loss):
\begin{equation}
\mathcal{L}_{MAE} = \frac{1}{N}\sum_{i=1}^{N} |y_i - \hat{y}_i|
\end{equation}

\subsubsection{Derivation: MAE as MLE under Laplace Noise}

If the noise follows a Laplace (double exponential) distribution:
\begin{equation}
\epsilon_i \sim \text{Laplace}(0, b), \quad p(\epsilon) = \frac{1}{2b}\exp\left(-\frac{|\epsilon|}{b}\right)
\end{equation}

Then:
\begin{align}
-\log p(y_i \mid \mathbf{x}_i; \theta) &= \log(2b) + \frac{|y_i - \hat{y}_i|}{b} \\
&\propto |y_i - \hat{y}_i|
\end{align}
Thus MAE is the MLE loss under Laplace noise. The Laplace distribution has heavier tails than the Gaussian, making MAE inherently more robust to outliers.

\subsubsection{Comparison: MSE vs MAE}

\begin{table}[ht]
    \centering
    \caption{MSE vs MAE: Comprehensive Comparison}
    \begin{tabular}{L{3cm} L{4cm} L{4.5cm}}
        \toprule
        \textbf{Property} & \textbf{MSE ($L_2$)} & \textbf{MAE ($L_1$)} \\
        \midrule
        Noise assumption & Gaussian $\mathcal{N}(0, \sigma^2)$ & Laplace $\text{Laplace}(0, b)$ \\
        Outlier sensitivity & High (quadratic penalty) & Low (linear penalty) \\
        Gradient w.r.t.\ $\hat{y}$ & $-2(y - \hat{y})/N$ & $-\text{sign}(y - \hat{y})/N$ \\
        At optimum $\hat{y}^*$ & Mean $\bar{y}$ & Median $\tilde{y}$ \\
        Smoothness & Smooth everywhere & Non-differentiable at 0 \\
        Convergence speed & Faster (quadratic near min) & Slower (constant gradient) \\
        Statistical efficiency & Higher (for Gaussian data) & Lower \\
        \bottomrule
    \end{tabular}
\end{table}

\textbf{Key insight:} The difference between MSE and MAE ultimately boils down to \emph{what you believe about the noise distribution}. If you expect occasional large outliers, MAE is more appropriate. If the errors are truly Gaussian, MSE is statistically optimal.

\subsection{Huber Loss: The Best of Both Worlds}

Huber loss \cite{huber1964robust} smoothly interpolates between MSE (for small errors) and MAE (for large errors):

\begin{equation}
\mathcal{L}_{\text{Huber}}(y, \hat{y}) = 
\begin{cases}
\frac{1}{2}(y - \hat{y})^2 & \text{if } |y - \hat{y}| \leq \delta \\
\delta \cdot |y - \hat{y}| - \frac{1}{2}\delta^2 & \text{otherwise}
\end{cases}
\end{equation}

The hyperparameter $\delta$ (typically $\delta = 1$) controls the transition point:
\begin{itemize}
    \item For $|e| < \delta$: Quadratic penalty (like MSE)---efficient optimization near the optimum.
    \item For $|e| \geq \delta$: Linear penalty (like MAE)---robust to outliers.
\end{itemize}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.9]
        \begin{axis}[
            xlabel={Error $e = y - \hat{y}$},
            ylabel=Loss,
            domain=-4:4,
            samples=200,
            width=11cm,
            height=6cm,
            legend pos=north west,
            ymin=0,
        ]
            \addplot[thick, blue] {x^2};
            \addlegendentry{MSE: $e^2$}
            \addplot[thick, red] {abs(x)};
            \addlegendentry{MAE: $|e|$}
            \addplot[thick, green, domain=-4:-1] {-2*x - 2};
            \addplot[thick, green, domain=-1:1] {x^2};
            \addplot[thick, green, domain=1:4] {2*x - 2};
            \addlegendentry{Huber: $\delta = 1$}
        \end{axis}
    \end{tikzpicture}
    \caption[Regression loss functions compared]{MSE, MAE, and Huber loss. Huber loss is quadratic near zero (fast convergence) and linear for large errors (outlier robustness).}
    \label{fig:regression_losses}
\end{figure}

\begin{example}[Huber Loss Gradient]
The gradient of Huber loss is continuous:
\begin{equation}
\frac{\partial \mathcal{L}_{\text{Huber}}}{\partial \hat{y}} = 
\begin{cases}
\hat{y} - y & \text{if } |y - \hat{y}| \leq \delta \\
\delta \cdot \text{sign}(\hat{y} - y) & \text{otherwise}
\end{cases}
\end{equation}
The gradient is clamped to $\pm \delta$ for large errors, preventing gradient explosion from outliers while maintaining smooth quadratic behavior near the optimum.
\end{example}

\section{Classification Loss Functions}

Classification tasks predict discrete class labels. The loss function must encourage the model to assign high probability to the correct class.

\subsection{From Likelihood to Cross-Entropy: A Complete Derivation Chain}

To truly understand why cross-entropy is the standard classification loss, we must trace the complete derivation from first principles.

\subsubsection{Step 1: The Generative Model}

For a $K$-class classification problem, we model the output as a categorical distribution parameterized by the network:
\begin{equation}
p(y = k \mid \mathbf{x}; \theta) = \hat{y}_k, \quad k = 1, \ldots, K
\end{equation}
where $\hat{y}_k = \text{softmax}(\mathbf{z})_k$ and $\mathbf{z} = f(\mathbf{x}; \theta)$ are the logits.

\subsubsection{Step 2: Maximum Likelihood}

Given $N$ i.i.d.\ samples, the likelihood is:
\begin{equation}
\mathcal{L}(\theta) = \prod_{i=1}^{N} p(y_i \mid \mathbf{x}_i; \theta) = \prod_{i=1}^{N} \prod_{k=1}^{K} \hat{y}_{ik}^{y_{ik}}
\end{equation}
where $y_{ik} \in \{0, 1\}$ is the one-hot encoding.

\subsubsection{Step 3: Negative Log-Likelihood = Cross-Entropy}

Taking the negative log:
\begin{align}
-\log \mathcal{L}(\theta) &= -\sum_{i=1}^{N} \sum_{k=1}^{K} y_{ik} \log \hat{y}_{ik} \\
&= -\sum_{i=1}^{N} \log \hat{y}_{i, c_i} \quad \text{(since } y_{ik} = \delta_{k, c_i}\text{)}
\end{align}
where $c_i$ is the true class for sample $i$. This is the \textbf{categorical cross-entropy loss}.

\subsubsection{Step 4: Connection to Information Theory}

Recall from Chapter 1 that the cross-entropy between the true distribution $p$ (one-hot: $p_{c_i} = 1$, $p_k = 0$ for $k \neq c_i$) and predicted distribution $q = \hat{\mathbf{y}}$ is:
\begin{align}
H(p, q) &= -\sum_{k=1}^{K} p_k \log q_k = -\log q_{c_i} \\
H(p) &= 0 \quad \text{(true distribution is deterministic)}
\end{align}
Therefore:
\begin{equation}
D_{KL}(p \| q) = H(p, q) - H(p) = -\log \hat{y}_{c_i} = \mathcal{L}_{\text{CE}}
\end{equation}
Minimizing cross-entropy $\equiv$ minimizing KL divergence $\equiv$ maximizing likelihood.

\subsubsection{Step 5: The Beautiful Gradient}

The softmax--cross-entropy gradient for a single sample:
\begin{align}
\frac{\partial \mathcal{L}}{\partial z_j} &= \frac{\partial}{\partial z_j}\left(-\log \frac{e^{z_{c}}}{\sum_k e^{z_k}}\right) \\
&= -\frac{\partial}{\partial z_j}\left(z_c - \log \sum_k e^{z_k}\right) \\
&= \begin{cases}
\hat{y}_j - 1 = \hat{y}_j - y_j & \text{if } j = c \\
\hat{y}_j - 0 = \hat{y}_j - y_j & \text{if } j \neq c
\end{cases} \\
&= \boxed{\hat{y}_j - y_j}
\end{align}

The gradient is simply \textbf{prediction minus truth}. This elegantly simple result means:
\begin{itemize}
    \item If $\hat{y}_c$ (predicted probability of correct class) is too low, the gradient pushes it up.
    \item If $\hat{y}_j$ (predicted probability of wrong class) is too high, the gradient pushes it down.
    \item The gradient automatically scales with confidence---a very wrong prediction produces a larger gradient.
\end{itemize}

\subsection{Binary Cross-Entropy}

For binary classification ($y \in \{0, 1\}$) with sigmoid output $\hat{y} = \sigma(z)$:
\begin{equation}
\mathcal{L}_{BCE} = -\frac{1}{N}\sum_{i=1}^{N}\left[y_i \log \hat{y}_i + (1-y_i) \log(1-\hat{y}_i)\right]
\end{equation}

This is a special case of categorical cross-entropy with $K = 2$ (and numerically more stable implementation). The gradient w.r.t.\ the logit $z$ is also elegantly simple:
\begin{equation}
\frac{\partial \mathcal{L}_{BCE}}{\partial z} = \hat{y} - y = \sigma(z) - y
\end{equation}

\subsection{Categorical Cross-Entropy}

For multi-class classification ($K$ classes):
\begin{equation}
\mathcal{L}_{CE} = -\frac{1}{N}\sum_{i=1}^{N}\sum_{k=1}^{K} y_{ik} \log \hat{y}_{ik}
\end{equation}

where $\hat{y}_{ik} = \text{softmax}(z_{ik})$.

\textbf{Numerical stability:} In practice, we compute the log-softmax directly:
\begin{equation}
\log \hat{y}_k = z_k - \log \sum_{j=1}^{K} e^{z_j} = z_k - \text{logsumexp}(\mathbf{z})
\end{equation}
The $\text{logsumexp}$ function is implemented using the log-sum-exp trick:
\begin{equation}
\text{logsumexp}(\mathbf{z}) = m + \log \sum_{j} e^{z_j - m}, \quad m = \max(\mathbf{z})
\end{equation}
This prevents overflow/underflow in the exponential.

\section{Advanced Loss Functions}

\subsection{Focal Loss}

Focal Loss \cite{lin2017focal} addresses the fundamental problem of \textbf{class imbalance} in single-stage detectors for dense object detection (RetinaNet). In object detection, the vast majority of proposed anchor boxes are background (easy negatives), so the standard cross-entropy loss is dominated by easily classified examples that provide little learning signal.

\subsubsection{Motivation: The Class Imbalance Problem}

Consider a one-stage object detector with $N$ anchors per image and typically only $O(1)$ positive (object) anchors. The cross-entropy loss:
\begin{equation}
\mathcal{L}_{CE} = -\log p_t
\end{equation}
where $p_t = p$ if $y = 1$ (positive) and $p_t = 1 - p$ if $y = 0$ (negative). The problem:
\begin{itemize}
    \item \textbf{Easy negatives} ($p_t \geq 0.9$): Even though there are millions, each contributes $-\log(0.9) \approx 0.105$ to the loss---small individually but enormous cumulatively.
    \item \textbf{Hard positives} ($p_t \leq 0.1$): Rare but critical, each contributing $-\log(0.1) \approx 2.3$---about 22$\times$ more than an easy negative.
    \item The combined effect: the gradient signal from hard examples is drowned out by the sheer number of easy examples.
\end{itemize}

\subsubsection{The Modulating Factor}

Focal Loss adds a modulating factor $(1 - p_t)^\gamma$ that down-weights the loss for well-classified examples:
\begin{equation}
\mathcal{L}_{\text{FL}} = -(1 - p_t)^\gamma \log p_t
\end{equation}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.9]
        \begin{axis}[
            xlabel={$p_t$ (probability of correct class)},
            ylabel=Loss,
            domain=0.01:0.99,
            samples=200,
            width=10cm,
            height=6cm,
            legend pos=north east,
            xmin=0, xmax=1,
            ymin=0,
        ]
            \addplot[thick, blue] {-ln(x)};
            \addlegendentry{CE ($\gamma=0$)}
            \addplot[thick, red] {(1-x)^1 * (-ln(x))};
            \addlegendentry{FL $\gamma=1$}
            \addplot[thick, green!70!black] {(1-x)^2 * (-ln(x))};
            \addlegendentry{FL $\gamma=2$}
            \addplot[thick, purple] {(1-x)^5 * (-ln(x))};
            \addlegendentry{FL $\gamma=5$}
        \end{axis}
    \end{tikzpicture}
    \caption[Focal loss modulation]{Focal Loss with different $\gamma$ values. As $\gamma$ increases, the loss for easy examples ($p_t \to 1$) is increasingly suppressed.}
    \label{fig:focal_loss}
\end{figure}

\subsubsection{Analysis of the Modulating Factor}

For different values of $p_t$ and $\gamma$:

\begin{table}[ht]
    \centering
    \caption[Focal Loss Modulating Factor]{Focal Loss Modulating Factor $(1 - p_t)^\gamma$}
    \begin{tabular}{L{2.5cm} c c c c}
        \toprule
        $p_t$ & $\gamma = 0$ & $\gamma = 1$ & $\gamma = 2$ & $\gamma = 5$ \\
        \midrule
        0.1 (hard) & 1.000 & 0.900 & 0.810 & 0.590 \\
        0.3 (medium) & 1.000 & 0.700 & 0.490 & 0.168 \\
        0.5 (ambiguous) & 1.000 & 0.500 & 0.250 & 0.031 \\
        0.9 (easy) & 1.000 & 0.100 & 0.010 & $\sim$0 \\
        0.99 (very easy) & 1.000 & 0.010 & 0.0001 & $\sim$0 \\
        \bottomrule
    \end{tabular}
\end{table}

Key observations:
\begin{itemize}
    \item When $\gamma = 0$, focal loss reduces to standard cross-entropy.
    \item For $\gamma = 2$: a $p_t = 0.9$ example contributes 100$\times$ less than a $p_t = 0.1$ example.
    \item For $\gamma = 5$: easy examples contribute essentially zero loss.
\end{itemize}

\subsubsection{Alpha-Balanced Focal Loss}

In practice, focal loss also incorporates a class-balancing weight $\alpha_t \in [0, 1]$:
\begin{equation}
\mathcal{L}_{\text{FL}} = -\alpha_t (1 - p_t)^\gamma \log p_t
\end{equation}

The interplay between $\alpha$ and $\gamma$ is subtle. In Lin et al.'s experiments, $\alpha = 0.25$ and $\gamma = 2$ achieved the best performance. The authors found that $\alpha$ matters less as $\gamma$ increases---because the modulating factor already effectively down-weights the majority class.

\subsubsection{Focal Loss Gradient}

For binary classification with $y \in \{0, 1\}$ and $p_t = p$ if $y = 1$, $p_t = 1 - p$ if $y = 0$:
\begin{equation}
\frac{\partial \mathcal{L}_{\text{FL}}}{\partial p} = \alpha_t \left[\gamma p_t^\gamma \log p_t - (1 - p_t)^{\gamma - 1}(1 - \gamma)\right] \cdot \frac{\partial p_t}{\partial p}
\end{equation}
The modulating factor $(1 - p_t)^\gamma$ naturally appears in the gradient, further emphasizing that the gradient magnitude for easy examples is suppressed.

\subsection{Label Smoothing}

Label smoothing \cite{szegedy2016rethinking}, introduced as part of the Inception-v2 architecture, replaces hard one-hot labels with soft labels. Despite its simplicity, it has become a standard technique in modern deep learning, especially for large language models (GPT, BERT).

\subsubsection{The Mechanism}

Given a hard label $\mathbf{y} = [0, \ldots, 1, \ldots, 0]$ (one-hot at position $c$), label smoothing constructs:
\begin{equation}
y_k^{\text{LS}} = 
\begin{cases}
1 - \epsilon + \frac{\epsilon}{K} & \text{if } k = c \\
\frac{\epsilon}{K} & \text{if } k \neq c
\end{cases}
\end{equation}
where $\epsilon \in [0, 1)$ is the smoothing parameter (typically $\epsilon = 0.1$). For example, with $K = 10$ classes and $\epsilon = 0.1$:
\begin{equation}
\mathbf{y}^{\text{LS}} = [0.01, 0.01, \ldots, 0.91, \ldots, 0.01]
\end{equation}

\subsubsection{Why Does Label Smoothing Work? Three Perspectives}

\textbf{Perspective 1: Preventing Overconfidence.}

Hard labels encourage the model to push logits to extreme values ($z_c \to +\infty$, $z_{k \neq c} \to -\infty$). This leads to:
\begin{itemize}
    \item \textbf{Large logit magnitudes}, increasing the softmax output gap.
    \item \textbf{Poor calibration}: The model's confidence does not match its accuracy.
    \item \textbf{Reduced generalization}: \cite{muller2019when} showed that a neural network with cross-entropy on hard labels has a minimum where $\|\mathbf{w}_c\| \gg \|\mathbf{w}_{k \neq c}\|$, i.e., the weight vector for the correct class grows much larger than others.
\end{itemize}

Label smoothing prevents this by assigning non-zero probability to all classes, bounding the maximum achievable confidence at $1 - \epsilon + \epsilon/K$.

\textbf{Perspective 2: Regularization via KL Divergence.}

The cross-entropy with smoothed labels decomposes as:
\begin{align}
\mathcal{L}_{\text{LS}} &= -\sum_k y_k^{\text{LS}} \log \hat{y}_k \\
&= -(1 - \epsilon)\underbrace{\log \hat{y}_c}_{\text{CE with hard label}} - \frac{\epsilon}{K}\underbrace{\sum_{k=1}^{K}\log \hat{y}_k}_{\text{negative entropy}} \\
&= (1 - \epsilon)\mathcal{L}_{CE} - \epsilon H(\hat{\mathbf{y}})
\end{align}

The second term $H(\hat{\mathbf{y}})$ is the \textbf{entropy of the predicted distribution}. Maximizing entropy encourages the predictions to be more uniform---a form of entropy regularization. This prevents the model from collapsing to very peaked distributions.

\textbf{Perspective 3: Bayesian Interpretation.}

Label smoothing can be viewed as incorporating a Dirichlet prior over the class probabilities. If $p \sim \text{Dir}(\alpha \mathbf{1})$ (symmetric Dirichlet), then:
\begin{equation}
\mathbb{E}[\log p_k] = \psi(\alpha) - \psi(K\alpha)
\end{equation}
where $\psi$ is the digamma function. For small $\alpha$, this approximates $\log \frac{1}{K}$, which corresponds to uniform label smoothing.

\subsubsection{Practical Impact}

Label smoothing has been shown to:
\begin{enumerate}
    \item Improve calibration (predicted probabilities better match true frequencies)
    \item Improve generalization on ImageNet (+0.2--0.5\% top-1 accuracy)
    \item Enable better knowledge distillation (the softer logits serve as better teacher signals)
    \item Reduce the vulnerability to adversarial examples
\end{enumerate}

\section{Contrastive and Self-Supervised Loss Functions}

Recent advances in self-supervised learning (2020--2024) have introduced a new family of loss functions that learn representations by contrasting positive pairs against negative pairs, without requiring manual labels.

\subsection{Contrastive Loss}

The simplest form of contrastive loss was introduced for metric learning (e.g., face verification, Siamese networks):

\begin{equation}
\mathcal{L}_{\text{contrastive}} = \frac{1}{2}\left[y \cdot d^2 + (1 - y) \cdot \max(0, m - d)^2\right]
\end{equation}
where:
\begin{itemize}
    \item $d = \|f(\mathbf{x}_1) - f(\mathbf{x}_2)\|_2$ is the distance between two embeddings
    \item $y \in \{0, 1\}$: $y = 1$ for a positive pair (same class), $y = 0$ for negative pair
    \item $m > 0$ is a margin hyperparameter
\end{itemize}

\textbf{Intuition:} Positive pairs are pushed close ($d \to 0$), while negative pairs are pushed apart until they are at least $m$ apart (beyond which no further penalty applies).

\subsection{Triplet Loss}

Triplet loss extends contrastive learning by comparing \textbf{anchor}, \textbf{positive}, and \textbf{negative} samples simultaneously:
\begin{equation}
\mathcal{L}_{\text{Triplet}} = \max\left(0, \underbrace{\|f_a - f_p\|_2^2}_{\text{positive distance}} - \underbrace{\|f_a - f_n\|_2^2}_{\text{negative distance}} + \alpha\right)
\end{equation}

\textbf{Geometric interpretation:} The loss is zero when the positive is at least $\alpha$ closer to the anchor than the negative. The margin $\alpha$ controls how much separation is enforced.

\subsection{InfoNCE Loss}

InfoNCE \cite{oord2018representation}, the foundation of modern contrastive learning (SimCLR \cite{chen2020simple}, MoCo, CLIP), frames representation learning as a \textbf{multiple-choice classification} problem:

\begin{definition}[InfoNCE Loss]
Given a query $\mathbf{q}$, one positive key $\mathbf{k}_+$, and $N-1$ negative keys $\{\mathbf{k}_-\}$:
\begin{equation}
\mathcal{L}_{\text{InfoNCE}} = -\log \frac{\exp(\mathbf{q}^\top \mathbf{k}_+ / \tau)}{\sum_{i=1}^{N} \exp(\mathbf{q}^\top \mathbf{k}_i / \tau)}
\end{equation}
where $\tau > 0$ is a temperature hyperparameter.
\end{definition}

\subsubsection{Connection to Mutual Information}

The remarkable theoretical result in \cite{oord2018representation} shows that InfoNCE provides a lower bound on mutual information:
\begin{equation}
I(\mathbf{x}; \mathbf{y}) \geq \log N - \mathcal{L}_{\text{InfoNCE}}
\end{equation}

This means minimizing InfoNCE maximizes the mutual information between the query and positive key, encouraging the encoder to capture all shared information between augmented views of the same image.

\subsubsection{The Role of Temperature \texorpdfstring{$\tau$}{tau}}

The temperature parameter $\tau$ controls the ``sharpness'' of the distribution:
\begin{itemize}
    \item \textbf{Small $\tau$} (e.g., 0.07 in CLIP): Makes the distribution sharp, increasing the penalty for negative samples that are too close. Encourages the model to be more ``discriminative.''
    \item \textbf{Large $\tau$} (e.g., 0.5--1.0): Softer distribution, more uniform gradients across negatives.
\end{itemize}

For a similarity score $s = \mathbf{q}^\top \mathbf{k} / \tau$ and $N$ negatives drawn from a uniform distribution over a hypersphere of dimension $d$:
\begin{equation}
\mathbb{E}[\mathcal{L}_{\text{InfoNCE}}] \approx -\log \sigma\!\left(\frac{s_{+} - \mathbb{E}[s_{-}]}{\sqrt{\text{Var}(s)}}\right) + \text{const}
\end{equation}
This shows that the loss effectively measures how separable the positive similarity is from the expected negative similarity.

\subsection{Dice Loss}

For segmentation tasks where class imbalance is severe (e.g., medical image segmentation), Dice loss \cite{milletari2016v} directly optimizes the Dice coefficient:
\begin{equation}
\mathcal{L}_{\text{Dice}} = 1 - \frac{2 \sum_{i} \hat{y}_i \cdot y_i + \epsilon}{\sum_{i} \hat{y}_i + \sum_{i} y_i + \epsilon}
\end{equation}

\subsubsection{Derivation: Dice Coefficient and Jaccard Index}

The \textbf{S{\o}rensen--Dice coefficient} measures the overlap between predicted and ground truth segmentation masks:
\begin{equation}
\text{Dice} = \frac{2|A \cap B|}{|A| + |B|}
\end{equation}

The \textbf{Jaccard index} (Intersection over Union, IoU) is related:
\begin{equation}
\text{IoU} = \frac{|A \cap B|}{|A \cup B|} = \frac{\text{Dice}}{2 - \text{Dice}}, \quad \text{Dice} = \frac{2 \cdot \text{IoU}}{1 + \text{IoU}}
\end{equation}

\textbf{Why Dice Loss works for class imbalance:} Unlike cross-entropy, which treats each pixel independently, Dice loss directly optimizes the overlap metric that is used for evaluation. This means:
\begin{itemize}
    \item A single false positive pixel reduces the numerator by 1 and increases the denominator by 1---a double penalty.
    \item The loss is inherently normalized by the size of both predicted and ground truth regions, making it scale-invariant.
    \item Small objects contribute proportionally more to the loss than they would under cross-entropy.
\end{itemize}

\textbf{Practical tip:} Dice loss is often combined with cross-entropy as $\mathcal{L} = \mathcal{L}_{CE} + \lambda \mathcal{L}_{\text{Dice}}$ with $\lambda \in [0.5, 1.0]$. Cross-entropy provides stable pixel-wise gradients, while Dice loss ensures region-level overlap.

\subsection{KL Divergence Loss}

The KL divergence loss is used when the target is itself a probability distribution (as opposed to a one-hot label):

\begin{equation}
\mathcal{L}_{KL}(p \| q) = \sum_{k} p_k \log \frac{p_k}{q_k}
\end{equation}

\textbf{Applications:}
\begin{itemize}
    \item \textbf{Knowledge distillation}: The target $p$ is the teacher's soft output distribution (after temperature scaling), and $q$ is the student's prediction.
    \item \textbf{Variational inference} (VAEs): Matching the approximate posterior $q(\mathbf{z}|\mathbf{x})$ to the prior $p(\mathbf{z})$.
    \item \textbf{Domain adaptation}: Aligning the feature distributions of source and target domains.
\end{itemize}

\textbf{Important: KL divergence is asymmetric.} $D_{KL}(p \| q) \neq D_{KL}(q \| p)$. In knowledge distillation:
\begin{itemize}
    \item \textbf{Forward KL} $D_{KL}(p_{\text{teacher}} \| q_{\text{student}})$: ``Mean-seeking''---the student covers all modes of the teacher. Used for distillation.
    \item \textbf{Reverse KL} $D_{KL}(q_{\text{student}} \| p_{\text{teacher}})$: ``Mode-seeking''---the student fits a few modes well but may miss others. Used in variational inference.
\end{itemize}

\section{Choosing a Loss Function}

\begin{table}[ht]
    \centering
    \caption{Loss Function Selection Guide}
    \begin{tabular}{L{2.5cm} L{4cm} L{4cm}}
        \toprule
        \textbf{Task} & \textbf{Loss} & \textbf{Notes} \\
        \midrule
        Binary Classification & BCE & Standard choice \\
        Multi-class & Cross-Entropy & With label smoothing \\
        Regression & MSE & Sensitive to outliers \\
        Robust Regression & MAE/Huber & Outlier resistant \\
        Imbalanced Classification & Focal Loss & Focus on hard examples \\
        Segmentation & Dice + CE & Handle class imbalance \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Chapter Summary}

Loss functions are the bridge between the mathematical model and the learning objective:

\begin{enumerate}
    \item \textbf{MSE = MLE under Gaussian noise}: Minimizing MSE is equivalent to fitting a model with additive Gaussian noise assumptions. The quadratic penalty makes it sensitive to outliers but efficient near the optimum.

    \item \textbf{MAE = MLE under Laplace noise}: The linear penalty provides robustness to outliers but yields slower convergence. Huber loss combines the advantages of both.

    \item \textbf{Cross-Entropy = MLE for categorical distributions}: The softmax--cross-entropy combination produces the elegant gradient $\hat{y}_k - y_k$, enabling efficient classification training.

    \item \textbf{Focal Loss}: Addresses class imbalance by dynamically down-weighting easy examples via a modulating factor $(1 - p_t)^\gamma$. Essential for dense prediction tasks.

    \item \textbf{Label Smoothing}: Prevents overconfidence by distributing probability mass across classes, with connections to Bayesian priors and entropy regularization.

    \item \textbf{Contrastive Losses (InfoNCE)}: The foundation of self-supervised learning, bounding mutual information between views of the same data point.

    \item \textbf{Dice Loss}: Directly optimizes the evaluation metric for segmentation, naturally handling class imbalance.
\end{enumerate}

\begin{table}[ht]
    \centering
    \caption{Complete Loss Function Decision Framework}
    \begin{tabular}{L{3cm} L{3cm} L{4cm} L{3cm}}
        \toprule
        \textbf{Scenario} & \textbf{Loss} & \textbf{Why} & \textbf{Key Ref} \\
        \midrule
        Clean regression & MSE & MLE, Gaussian noise & \\
        Regression + outliers & Huber ($\delta=1$) & Robust + efficient & \cite{huber1964robust} \\
        Binary classification & BCE & MLE, Bernoulli & \\
        Multi-class & Cross-Entropy & MLE, categorical & \\
        Imbalanced detection & Focal Loss ($\gamma=2$) & Down-weights easy negs & \cite{lin2017focal} \\
        Medical segmentation & Dice + CE & Overlap optimization & \cite{milletari2016v} \\
        Self-supervised & InfoNCE ($\tau=0.07$) & MI maximization & \cite{chen2020simple} \\
        Knowledge distillation & KL Divergence & Distribution matching & \\
        NLP / large models & CE + Label Smooth & Calibration & \cite{szegedy2016rethinking} \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Further Reading}

\begin{itemize}
    \item \cite{huber1964robust} --- Robust Estimation of a Location Parameter (Huber Loss)
    \item \cite{bishop2006pattern} --- Chapters 1--2 for probabilistic foundations
    \item \cite{goodfellow2016deep} --- Chapter 5 (Machine Learning Basics) and Chapter 6 (Deep Feedforward Networks)
    \item \cite{milletari2016v} --- V-Net: Dice Loss for segmentation
    \item \cite{szegedy2016rethinking} --- Rethinking the Inception Architecture (Label Smoothing)
    \item \cite{lin2017focal} --- Focal Loss for Dense Object Detection
    \item \cite{oord2018representation} --- Contrastive Predictive Coding (InfoNCE)
    \item \cite{chen2020simple} --- SimCLR: Simple Framework for Contrastive Learning
    \item \cite{he2020masked} --- Masked Autoencoders (MAE) for self-supervised vision
\end{itemize}

\section{Exercises}

\begin{enumerate}
    \item \textbf{Huber vs. MSE:} For residuals $r \in \{-3, -1, 0, 1, 3\}$ and Huber threshold $\delta = 1$, compute the Huber loss for each residual and compare it with MSE and MAE. Which loss penalizes large outliers most strongly?

    \item \textbf{Softmax-Cross-Entropy Gradient:} Starting from $\mathcal{L} = -\sum_k y_k \log \hat{y}_k$ with $\hat{y}_k = \text{softmax}(z_k)$, derive the gradient $\frac{\partial \mathcal{L}}{\partial z_k} = \hat{y}_k - y_k$.

    \item \textbf{Class Imbalance:} Consider a binary dataset with 95\% negatives and 5\% positives. Explain why standard BCE may underperform. How does focal loss change the contribution of easy negatives during training?

    \item \textbf{Label Smoothing:} For a 5-class classification task with smoothing parameter $\varepsilon = 0.1$, write the smoothed target distribution for the correct class and explain why this can improve calibration.

    \item \textbf{Contrastive Learning:} In InfoNCE with batch size $N$, each sample has one positive and $2N-2$ negatives. How does increasing batch size affect the denominator of the objective and the difficulty of the task?

    \item \textbf{Segmentation Losses:} Show how Dice loss changes when the predicted mask over-segments the object compared with when it under-segments it. Why is Dice loss often combined with cross-entropy in practice?
\end{enumerate}

